paper2_resistance_mechanisms_mixtures

### 2016-12-15 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
rewriting results

---
title: "paper2_resistance_mechanisms_mixtures"
output: pdf_document
#output: word_document
author: "Andy South and Ian Hastings"
date: "`r Sys.Date()`"
bibliography: library_ir.bib
csl: biomed-central.csl
---

```{r,echo=FALSE,results='hide'}
  library(resistance)
  library(ggplot2)
```

@@

## ANDY TO DO
1. add to intro including about about results in levick et al.
1. work out what's going on with exposure ... why does increasing exposure have a greater effect on mixture than sequential ? 
1. explain resistance-restoration and its relation to selection coefficient
1. expand discussion to include relevance to real world 
1. in discussion address potential criticism that these results are just for a limited region of paramter space (a. point to Levick et al. that explored more of space, b. point to UI that allows readers to explore spce for themselves and identify whether there are places where this does not hold).
1. ? briefly refer to linkage disequilibrium and say we are investigating.

@@

Insecticide resistance has been identified as a major problem for the control of malarial mosquitoes with potential huge consequences for malaria burden. To address this problem effort is being made to develop new insecticide tools including new active ingredients and combinations with existing ones. There are different potential strategies for using new insecticides and recomendations for how best to preserve their utility for controlling mosquitoes and malaria. There is however limited recent, accessible modelling work to support suggestions of which strategies may be best in different situations. Previous work suggests that preferred strategies will be situation specific. This points to a need for an accessible, mechanistic understanding of how the evolution of insecticide resistance is likely to respond to potential intervention strategies to help guide both management and policy. Such a mechanistic understanding would be more robust to uncertainty in parameter values in field situations.

@@

Insecticide resistance becomes a problem when genes coding for resistance firstly arise in a population and secondly increase in frequency. We concern ourselves with this second process of how insecticide resistance increases in frequency within a population. _is their research indicating that mnost mutations already there? something from ASTMH?_

@@

So far this description has just considered a single insecticide and associated resistance allele (Fig 1A). In the simulation a second insecticide and resistance allele are included and fitness is calculated by multiplying these with those for the first insecticide described above (Fig 1C). Thus the simulation can represent populations exposed to two insecticides together in a mixture.

In addition the model structure allows for 3 different levels of exposure to be set for each insecticide (Fig 2). So in the two insecticide case that gives the potential for nine (3*3) different exposure niches. In the runs shown here we have just used two levels of exposure for each insecticide (effectively no or yes with the value for yes set from the exposure input in Table 1). The model structure allows males and females to experience different exposures, we explore that in Levick et al. (in press) but the runs presented here use equal exposures.

@@

### Single insecticide
For single insecticide use, faster resistance spread resulted from higher values of any of the inputs outlined in Table 1, namely effectiveness, exposure, dominance, the resistance restoration coefficient, and starting frequencies of resistance (Figs 3-5). In the plots each input was varied in isolation with the remaining set to 0.5 (except for starting frequency of resistance which was set to 0.01 in all bar Fig 5).

### Two insecticides
When two insecticides are used the relative performance of a mixture strategy compared to sequential use changes dependent on input values. To help us understand the mechanisms influencing this relative performance we started with a single base scenario and investigated changing inputs individually. The base scenario was similar to that used for the single insecticide. Values for the first four inputs in Table 1 were set to 0.5 and the starting frequencies of resistance to 0.01. For this base scenario resistance arises slower for sequential than mixture (Fig 6A). This plot, and those that follow, compares mixture and sequence for a single combination of inputs. Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time (as would be expected given that they have identical input parameters). When used in sequence the curve (dashed) for each insecticide individually is steeper than in the mixture but, because they happen one after the other, it takes longer for both to reach the resistance threshold of 50%.

_TODO rewrite this & probably put effectiveness and exposure in together_
_starting like this_
In the baseline scenario where the effectiveness of both insecticides is set the same at 0.5 the resistance frequencies at each locus for the mixture remain identical (the single solid line in Fig 6A). For this scenario time-to-resistance is longer when the insecticides are used in sequence (dashed lines). Increasing exposure to the insecticides from 0.5 to 0.8 (Fig 6C) decreases the time-to-resistance for both the sequential and mixture strategy and so does not change their relative performance; time-to-resistance remains longer for the sequence.

#### results: Insecticide effectiveness
A more effective insecticide increases time-to-resistance when used in a mixture. This is opposite to what happens when used in sequence and in isolation, when both shortened times-to-resistance.

For example, taking the base scenario and increasing the effectiveness of one insecticide from 0.5 to 0.8 increases the time-to-resistance for the mixture (from 90 to 115 generations) and decreases that for the sequence (from 120 to 90 generations) (Fig 6). Thus at low effectiveness a sequential strategy is favoured while at higher effectiveness a mixture is favoured.

When the effectiveness of both insecticides are increased to 0.8 the time-to-resistance is increased still further to around 125 generations (Fig 7).

#### results: Exposure
Increasing exposure to 0.8 from the base scenario reduces time-to-resistance for both mixture and sequential scenarios (Fig 8A relative to 6A). The increase in exposure has a greater effect on the mixture relative to the sequential strategy and this results in a greater relative benefit for sequence versus mixture.

Similarly increasing exposure reduces time-to-resistance more for the mixture relative to sequence in the scenario in which one insecticide effectiveness has been increased (compare Fig 8B to 6B). The result is that the advantage of mixtures over sequence is removed by the increase in exposure.

#### results: Dominance of resistance alleles
Increasing the dominance of the alleles within a mixture has the same direction of effect as it does for sole use. Increasing the dominance of the allele corresponding to insecticide1 from 0.5 to 0.8 decreases the time-to-resistance to insecticide1 within the mixture but leaves the time-to-resistance for the second insecticide in the mixture the same (Fig 9A). Setting the dominance for both alleles to 0.8 reduces the time-to-resistance for both the mixture and the sequence by 20 generations from the base scenario (compare Fig 9B to 6A). In both of these cases sequential use continues to have a longer time-to-resistance under the higher dominance.

#### results: Resistance restoration coefficient
Increasing the resistance restoration coefficient to insecticide1 from 0.5 to 0.8 decreases the time-to-resistance for insecticide1 in the mixture, and has a similar effect on it's use in the sequence (Fig 10A). However, it has little effect on resistance to insecticide2. Because of this sequential use continues to generate a longer time-to-resistance than the mixture. Increasing the resistance restoration coefficient of both alleles to 0.8 results in a shorter time-to-resistance  for the mixture (Fig 10B), however the relative shape of the resistance curves remains similar to the base scenario  (Fig 6A), with sequential use continuing to generate longer time-to-resistance.

#### results: Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 11A to Fig 6A). Similarly taking the base scenario with effectiveness of the first insecticide increased to 0.8 (Fig 6B) and decreasing starting frequency of that resistance allele did not result in the mixture being favoured (Fig 11B).

@@

When single insecticide use was represented (Figs 3-5) in all cases the initial increase in resistance was most rapid with straight lines on the log plots up to around 50% followed by a slowing rate of increase up to 100%. Higher values of the five inputs investigated all led to more rapid increases in resistance.

For single insecticide use resistance responded identically to changing exposure and effectiveness (compare Fig 3A to 3B). This makes sense as vector kill is effectively a product of exposure times effectiveness. For example exposing 50% of a  population to an insecticide which is 75% effective would be expected to have the same result as exposing 75% of the population to one which is 50% effective. This observation points to the mechanism by which increasing both exposure and effectiveness lead to a faster increase in resistance. In both cases the increased deaths of susceptible vectors cause a higher selection pressure that can explain the faster increase.

The pattern of more rapid increase in resistance at higher dominance levels (Fig 4A) can be explained by increased survival of heterozygotes in the presence of the insecticide. Higher dominance of the resistant allele causes it to contribute more to the phenotype of the heterozygotes leading to higher survival. Thus selection pressure for the resistance allele will be increased because it confers more of an advantage when only present on one chromosome.

The faster development of resistance under higher resistance-restoration (Fig 4B) can be explained by it's effect on the survival of the resistant genotypes. Resistance-restoration restores the survival of resistant genotypes in the presence of the insecticide back towards what it would be in the absence of the insecticide, thus increasing the selection pressure of the resistance allele.

The effect of the starting frequency of resistance (Fig 5) is the most different from the other inputs and perhaps the easiest to explain. Simply when starting from a higher frequency of resistance there is a smaller change to make to reach the higher resistance thresholds.

These mechanisms explaining the response of resistance to single insecticide use under different input values are summarised in Table 2 and Fig 12. Exposure and effectiveness act by decreasing the survival of all genotypes, dominance acts by increasing survival of RS and resistance-restoration acts by increasing survival of RR & RS.

@@

In comparison with the use of a single insecticide, using two insecticides either in sequence or in mixtures differs in some ways and where it differs that can again help us to understand the mechanisms. Looking at our base scenario (Fig 6A) the curves for resistance to both insecticides in the mixture are shallower than those for sequential use and in this particular example are identical because they have set the same parameter values for each insecticide to be the same. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Thus each insecticide reduces the selection pressure for the increase in resistance to the other and could be said to 'protect' the other. However this protection is not necessarily enough to ensure that resistance arises more slowly in a mixture relative to a sequential strategy. As shown in our base scenario, resistance can arise more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other.

#### discussion: Insecticide effectiveness

The main difference between mixtures and the other strategies is that in mixtures increasing the effectiveness of either insecticide slows the development of resistance to both (Fig 6B). In contrast in both the sequential (Fig 6B) and single insecticide (Fig 3A) experiments resistance was speeded up by increasing effectiveness. This can be explained by the effect that in a mixture increased effectiveness of one insecticide increases killing of vectors that are resistant to the other and thus decreases selection pressure.

This effect can be seen by comparing Fig 6B, where the effectiveness of insecticide 1 has been increased, to 6A. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. This can be explained by the mechanism that once insecticide 1 has become ineffective due to resistance it stops killing indivduals that are resistant to insecticide 2 and thus stops 'protecting' it from it's own rise in resistance. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 7). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 6A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 6B.

Fig 6B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal resistance will be expected to increase faster to the more effective insecticide. The more effective insecticide prompts both a) higher selection pressure for it's own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide.

@@

![Exposure to insecticides in the model.](paper2_mod_graphic_d.png)

@@

![Summarising mechanisms of single insecticide effects.](paper2_input_effects_single.png)

>>>>

---
title: "paper2_resistance_mechanisms_mixtures"
output: pdf_document
#output: word_document
author: "Andy South and Ian Hastings"
date: "`r Sys.Date()`"
bibliography: IR.bib
csl: biomed-central.csl
---

```{r,echo=FALSE,results='hide',warning=FALSE}
  library(resistance)
  library(ggplot2)
```

@@

## ANDY TO DO
1. add a screenshot of the UI
1. add to intro including about about results in levick et al.
1. work out what's going on with exposure ... why does increasing exposure have a greater effect on mixture than sequential ? 
1. explain resistance-restoration and its relation to selection coefficient
1. expand discussion to include relevance to real world 
1. in discussion address potential criticism that these results are just for a limited region of paramter space (a. point to Levick et al. that explored more of space, b. point to UI that allows readers to explore spce for themselves and identify whether there are places where this does not hold).
1. ? briefly refer to linkage disequilibrium and say we are investigating.

@@

Insecticide resistance has been identified as a major problem for the control of malarial mosquitoes with potentially huge consequences for malaria burden. To address this problem effort is being made to develop new insecticide tools including new active ingredients and combinations with existing ones. There are different potential strategies for using new insecticides and recomendations for how best to preserve their utility for controlling mosquitoes and malaria. There is however limited recent, accessible modelling work to support suggestions of which strategies may be best in different situations. Previous work suggests that preferred strategies will be situation specific. This points to a need for an accessible, mechanistic understanding of how the evolution of insecticide resistance is likely to respond to potential intervention strategies to help guide both management and policy. Such a mechanistic understanding would be more robust to uncertainty in parameter values in field situations.

@@

Insecticide resistance becomes a problem when genes coding for resistance firstly arise in a population and secondly increase in frequency. We concern ourselves with this second process of how insecticide resistance increases in frequency within a population. _is their research indicating that most mutations already there? something from ASTMH?_

@@

So far this description just considers a single insecticide and associated resistance allele, with equal behaviour for males and females (Fig 1A). To allow males and females to be exposed differently separate genders are also represented (Fig 1B). In addition a second insecticide and resistance allele are included and fitness is calculated by multiplying these with those for the first insecticide described above (Fig 1C). Thus the simulation can represent populations exposed to two insecticides together in a mixture.

@@

### Single insecticide
For single insecticide use, faster resistance spread resulted from higher values of any of the inputs outlined in Table 1, namely effectiveness, exposure, dominance, the resistance restoration coefficient, and starting frequencies of resistance (Figs 2-4). In the plots each input was varied in isolation with the remaining set to 0.5 (except for starting frequency of resistance which was set to 0.01 in all bar Fig 4).

### Two insecticides
When two insecticides are used the relative performance of a mixture strategy compared to sequential use changes dependent on input values. To help us understand the mechanisms influencing this relative performance we started with a single base scenario and investigated changing inputs individually. The base scenario was similar to that used for the single insecticide. Values for the first four inputs in Table 1 were set to 0.5 and the starting frequencies of resistance to 0.01. For this base scenario resistance arises slower for sequential than mixture (Fig 5A). This plot, and those that follow, compares mixture and sequence for a single combination of inputs. Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time (as would be expected given that they have identical input parameters). When used in sequence the curve (dashed) for each insecticide individually is steeper than in the mixture but, because they happen one after the other, it takes longer for both to reach the resistance threshold of 50%.

#### results: Insecticide exposure and effectiveness

Increasing exposure to the insecticides (the proportion of insects that come into contact) from 0.5 to 0.8 (Fig 5C) decreases the time-to-resistance for both the sequential and mixture strategy and so does not change their relative performance; time-to-resistance remains longer for the sequence. In contrast keeping the exposure constant at 0.5 and increasing effectiveness (the proportion of SS insects that are killed by contact) of one of the insecticides to 0.8 (Fig 5D) results in a longer time-to-resistance for the mixture relative to the sequence. Resistance for the more effective insecticide (in red) increases faster than the lesser. Resistance to the less effective insecticide in the mixture increases slowly initially and then speeds up after the more effecttive insecticide has reached the resistance threshold (~50 generations).

Thus a more effective insecticide increases time-to-resistance when used in a mixture (compare the solid lines in Fig 5A & B). This is opposite to what happens when used in sequence (compare the red dashed line in Fig 5A & B) and in isolation (Fig 2A), when more effective insecticides shortened times-to-resistance.

Figure 5D has the same inputs as Fig 5B except that the exposure is increased from 0.5 to 0.8. The increase in exposure reduces time-to-resistance more for the mixture relative to the sequence. The result is that the advantage of mixtures over sequence is removed by the increase in exposure, leading to approximately equal times-to-resistance for the two strategies (Fig 5D).

Figure 6 shows how increasing the effectiveness of either or both insecticides improves the performance of mixtures over sequences in terms of time-to-resistance. Figure 6A uses the same inputs as Fig 5B. From this scenario a higher effectiveness for insecticide2 (Fig 6C) or insecticide 1 (Fig 6B) or both (Fig 6D) all result in a greater positive difference in time-to-resistance for mixture over sequence.

#### results: Dominance and resistance restoration
Increasing dominance or resistance restoration decreases time-to-resistance for both sequences and mixtures (Fig 7) in the same way that it did for sole use (Fig 3). The result is that increasing either or both of dominance and resistance restoration (Figs 7B,C,D) doesn't change the relative time-to-resistance for mixtures and sequences from that in the base scenario (Fig 7A) where it takes resistance longer to develop for the sequence.

#### results: Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 8C to 8A). Similarly taking a scenario in which time-to-resistance is longer for a mixture (Fig 8B) and decreasing the starting frequency of that resistance allele did not change the fact that the mixture was favoured (Fig 8D).

TODO Fig 9 male exposure proportion

@@

When single insecticide use was represented (Figs 2-4) in all cases the initial increase in resistance was most rapid with straight lines on the log plots up to around 50% followed by a slowing rate of increase up to 100%. Higher values of the five inputs investigated all led to more rapid increases in resistance.

For single insecticide use resistance responded identically to changing exposure and effectiveness (compare Fig 2A to 2B). This makes sense as vector kill is effectively a product of exposure times effectiveness. For example exposing 50% of a  population to an insecticide which is 75% effective would be expected to have the same result as exposing 75% of the population to one which is 50% effective. This observation points to the mechanism by which increasing both exposure and effectiveness lead to a faster increase in resistance. In both cases the increased deaths of susceptible vectors cause a higher selection pressure that can explain the faster increase.

The pattern of more rapid increase in resistance at higher dominance levels (Fig 3A) can be explained by increased survival of heterozygotes in the presence of the insecticide. Higher dominance of the resistant allele causes it to contribute more to the phenotype of the heterozygotes leading to higher survival. Thus selection pressure for the resistance allele will be increased because it confers more of an advantage when only present on one chromosome.

The faster development of resistance under higher resistance-restoration (Fig 3B) can be explained by it's effect on the survival of the resistant genotypes. Resistance-restoration restores the survival of resistant genotypes in the presence of the insecticide back towards what it would be in the absence of the insecticide, thus increasing the selection pressure of the resistance allele.

The effect of the starting frequency of resistance (Fig 4) is the most different from the other inputs and perhaps the easiest to explain. Simply when starting from a higher frequency of resistance there is a smaller change to make to reach the higher resistance thresholds.

These mechanisms explaining the response of resistance to single insecticide use under different input values are summarised in Table 2. Exposure and effectiveness act by decreasing the survival of all genotypes, dominance acts by increasing survival of RS and resistance-restoration acts by increasing survival of RR & RS.

@@

In comparison with the use of a single insecticide, using two insecticides either in sequence or in mixtures differs in some ways and where it differs that can again help us to understand the mechanisms. Looking at our base scenario (Fig 5A) the curves for resistance to both insecticides in the mixture are shallower than those for sequential use and in this particular example are identical because they have set the same parameter values for each insecticide to be the same. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Thus each insecticide reduces the selection pressure for the increase in resistance to the other and could be said to 'protect' the other. However this protection is not necessarily enough to ensure that resistance arises more slowly in a mixture relative to a sequential strategy. As shown in our base scenario, resistance can arise more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other.

TODO check figure numbering in next section
#### discussion: Insecticide effectiveness

The main difference between mixtures and the other strategies is that in mixtures increasing the effectiveness of either insecticide slows the development of resistance to both (Fig 5B). In contrast in both the sequential (Fig 5B) and single insecticide (Fig 2A) experiments resistance was speeded up by increasing effectiveness. This can be explained by the effect that in a mixture increased effectiveness of one insecticide increases killing of vectors that are resistant to the other and thus decreases selection pressure.

This effect can be seen by comparing Fig 5B, where the effectiveness of insecticide 1 has been increased, to 5A. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. This can be explained by the mechanism that once insecticide 1 has become ineffective due to resistance it stops killing indivduals that are resistant to insecticide 2 and thus stops 'protecting' it from it's own rise in resistance. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 6C). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 5A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 5B.

Fig 5B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal resistance will be expected to increase faster to the more effective insecticide. The more effective insecticide prompts both a) higher selection pressure for it's own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide.

@@

@@
